Vagueness, Noise, and Communication

### 2014-10-02 paper/01_intro.tex
discussion section; first almost complete version

More concretely, a sim-max game, in the relevant sense here, consists
of a set of states $\States$, a set of messages $\Messgs$ with much
fever messages than states, a probability distribution $\Pr \in
\Delta(\States)$ that gives the occurrence probabilities of states, a
similarity metric on states $\similarity \mycolon \States \times
\States \rightarrow \mathds{R}$, and a utility function $\Utils
\mycolon \States \times \States \rightarrow \mathds{R}$. We identify
the receiver's acts with the states of the world, so that the game is
one of guessing the actual state, so to speak. We also assume that
sender's and receiver's interests are alike, so we only have one
utility function. We do not consider message costs, so utilities only
depend on the actual state, and the receiver's response. The
similarity function should satisfy the usual requirements for a
metric:
\begin{align*}
  & \similarity(\mystate{1}, \mystate{2}) \ge 0 &&
  \similarity(\mystate{1}, \mystate{2}) = 0 \Leftrightarrow
  \mystate{1} = \mystate{2} \\
  & \similarity(\mystate{1}, \mystate{2}) = \similarity(\mystate{2},
  \mystate{1}) && \similarity(\mystate{1}, \mystate{2}) \le
  \similarity(\mystate{1}, \mystate{3}) + \similarity(\mystate{3}, \mystate{2})\,.
\end{align*}
The utility function should be a monotonically decreasing function of
similarity:
\begin{align*}
  \similarity(\mystate{1},\mystate{2}) \ge
  \similarity(\mystate{1},\mystate{3}) \ \ \Rightarrow \ \ 
  \Utils(\mystate{1},\mystate{2}) \ge
  \Utils(\mystate{1},\mystate{3})\,.
\end{align*}
To keep matters simple, we only look at cases where $\States$ contains
finitely many points from the unit interval, all of which have equal
probability, and with similarity given by the Euclidean distance in
this one-dimensional space.\todo{introduce utilities here?}

>>>>

More concretely, a sim-max game, in the relevant sense here, consists
of a set of states $\States$, a set of messages $\Messgs$ with much
fever messages than states, a probability distribution $\Pr \in
\Delta(\States)$ that gives the occurrence probabilities of states, a
similarity metric on states $\similarity \mycolon \States \times
\States \rightarrow \mathds{R}$, and a utility function $\Utils
\mycolon \States \times \States \rightarrow \mathds{R}$. We identify
the receiver's acts with the states of the world, so that the game is
one of guessing the actual state, so to speak. We also assume that
sender's and receiver's interests are alike, so we only have one
utility function. We do not consider message costs, so utilities only
depend on the actual state, and the receiver's response. The
similarity function should satisfy the usual requirements for a
metric:
\begin{align*}
  & \similarity(\mystate{1}, \mystate{2}) \ge 0 &&
  \similarity(\mystate{1}, \mystate{2}) = 0 \Leftrightarrow
  \mystate{1} = \mystate{2} \\
  & \similarity(\mystate{1}, \mystate{2}) = \similarity(\mystate{2},
  \mystate{1}) && \similarity(\mystate{1}, \mystate{2}) \le
  \similarity(\mystate{1}, \mystate{3}) + \similarity(\mystate{3}, \mystate{2})\,.
\end{align*}
The utility function should be a monotonically decreasing function of
similarity:
\begin{align*}
  \similarity(\mystate{1},\mystate{2}) \ge
  \similarity(\mystate{1},\mystate{3}) \ \ \Rightarrow \ \ 
  \Utils(\mystate{1},\mystate{2}) \ge
  \Utils(\mystate{1},\mystate{3})\,.
\end{align*}
To keep matters simple, we only look at cases where $\States$ contains
finitely many points from the unit interval, all of which have equal
probability, and with similarity given by the Euclidean distance in
this one-dimensional space.

### 2014-10-02 paper/02_noisyRMD.tex
introduction and background; sec com
discussion section; first almost complete version

\section{Replicator diffusion dynamic}

@@

\subsection{Replicator diffusion dynamic in behavioral strategies}

@@

The idea behind the sequential definition of the \rdd is that, at each
time step, strategies are gradually optimized along the current
fitness landscape, as described by the \rd, but the realization of
optimized strategies is noisy, due to confusion of states. It is
obvious that other (discrete-time) evolutionary dynamics can be
subjected to state-confusability in an analogous fashion. In the case
of the \rd, however, perturbation by state-confusion has a prominent
close relative in the replicator mutator dynamic.

>>>>

\section{Replicator diffusion dynamic}
\label{sec:repl-diff-dynam}

@@

\subsection{Replicator diffusion dynamic in behavioral strategies}
\label{sec:repl-diff-dynam-1}

@@

The idea behind the sequential definition of the \rdd is that, at each
time step, strategies are gradually optimized along the current
fitness landscape, as described by the \rd, but the realization of
optimized strategies is noisy, due to confusion of states. It is
obvious that other (discrete-time) evolutionary dynamics can be
subjected to state-confusability in an analogous fashion (see also
Section~\ref{sec:model-interpretation}). In the case of the \rd,
however, perturbation by state-confusion has a prominent close
relative in the replicator mutator dynamic.

### 2014-10-02 paper/04_discussion.tex
discussion section; first almost complete version

\subsection{Model interpretation}
The replicator-mutator equations are a mathematical model motivated by concepts from the theory of natural selection, namely natural variation, differential reproduction, and mutation.
The most straightforward interpretation is thus a biological one: we can imagine signaling strategies as innate behaviorial tendencies of organisms, steps in the evolutionary process as successive generations, and selection as capturing the reproductive advantage of fitter individuals.
In this interpretation, our proposed model is not an extremely interesting one, since the noise matrix simply corresponds to a restricted form of mutation.

Another possibility is to take the equations as embodying cultural evolution.
In this interpretation, what is subject to the evolutionary forces are not organisms but strategies themselves.
This pressuposes that individuals can adopt strategies from other individuals.
Fitness captures the success of strategies, which in the case of language can be thought of in terms of communicative success.
Differential reproduction represents the tendency of more successful strategies to be more likely adopted by individuals, be it through imitation or some learning procedure within the population.
Under this interpretation, the proposed noise perturbation becomes more interesting.
Messages correspond to the overt data available to individuals, whereas states are interpreted as the more private aspect of the behavior.
Futhermore, similarity-maximization games are especially interesting in situations where the number of states largely exceeds the number of messages available.
When trying to adopt another individual's behavior, it is thus more likely that information about states is both less accessible and less practical to elicit, hence justifying a higher likelihood of mistakes and a more pressing need for a certain degree of generalization%
\footnote{The consequence of perturbing a strategy with a noise matrix based on similarity can be seen as ``spreading'' the effects of the fitness of the behavior for a state to its more similar states, which can also be interpreted as generalization}%
.

\subsection{Relation with previous accounts}

@@

\paragraph{The conceptual spaces approach.}

\begin{itemize}
\item can only ``construct'' (fuzzy) tessellation's given the
  prototypes (or prototype regions); but cannot explain how the
  prototypes come about
\end{itemize}

>>>>

We have introduced a new variant of the replicator mutator dynamics
that implements specifically stochastic noise in the form of
confusability of similar states. The probabilities of confusing nearby
similar states is easily implemented in behavioral strategies, but can
also be translated into a mutation matrix for pure strategies. This
allows, in principle, several conceptual interpretations of the \rdd,
on which we will briefly enlarge below in
Section~\ref{sec:model-interpretation}.

Next to this technical contribution, the results of our numerical
simulations also advance our understanding of the possibility of
evolving regular and efficient categories despite their (higher-order)
vagueness. Section~\ref{sec:relat-with-prev} zooms in on the relation
with closely related accounts once more, to delineate the present
approach more precisely.

\subsection{Model interpretation}
\label{sec:model-interpretation}

We mentioned in passing in Section~\ref{sec:repl-diff-dynam-1} that
adding diffusion to other discrete-time evolutionary dynamics that
operate on behavioral strategies is entirely straightforward. We
could, for instance, easily diffuse the outcome of a best-response
dynamic at each time step. That would make sense if we thought of
agents as prone to confusing similar states that are otherwise
rational optimizers of behavior. The reason that we chose the
replicator dynamics to combine with diffusion is twofold. A minor
reason is that it makes for a conceptually interesting link with the
replicator mutator dynamics. A more important reason is that the
replicator dynamic is especially versatile and non-committal about
what the exact process of adaption is that is being modelled.

Originally the \rd was introduced as mathematical model of evolution
under asexual reproduction, motivated by concepts from the theory of
natural selection \citep{TaylorJonker1978:Evolutionary-St}. The most
conservative interpretation of the \rdd in our present context is thus
a biological one: we can imagine signaling strategies as innate or
fixed behaviorial tendencies of organisms, steps in the evolutionary
process as successive generations, and selection as capturing the
reproductive advantage of fitter individuals. This interpretation,
strictly speaking, requires formalization in terms of mixed strategies
via the \rmd, and possibly also a symmetrizing of the game, so that
every agent is assumed to have a unique sender and receiver role at
the same time the is bequeathed onto the next generation. Diffusion,
in this context, could be either of two things. It could be
differential mutation probabilities in line with other interpretation
of mutation in the \rmd
\citep[e.g.][]{NowakKomarova2001:Evolution-of-Un,KomarovaNiyogi2001:The-Evolutionar}:
pure strategies are not faithfully reproduced, and mistakes in
bequeathing pure strategies are more likely if they result from the
confusion of similar states. Another possible biological
interpretation of the \rdd, is that inheritance is faithful, but
strategies are noisily realized. Strategies are not necessarily
selected for what they are, but rather for how they are realized, once
noise is factored in.

But the replicator dynamic is not only a model of biological
evolution. It can also be interpreted as a high-level description of
the likely development of other behavioral adaptation processes, like
differential imitation and cultural evolution in general
\citep[see][for various derivations of the
\rd]{Sandholm2013:Population-Game}. Under this interpretation, the \rd
is consistent with the idea that what is subject to the evolutionary
forces are not organisms but behavior: individuals can adapt their
behavior to the perceived environment or adopt strategies from other
individuals. Fitness captures the success of behavioral patterns,
which in the case of language can be thought of in terms of
communicative success.  Differential reproduction represents the
tendency of more successful strategies to be more likely adopted by
individuals, be it through imitation or some learning procedure within
the population.

Under this non-biological interpretation, we have again two options of
picturing what diffusion is, similar to the biological cases
before. One possibility is that diffusion is noise in the adoption of
strategies, say, by conditional imitation, of strategies by other
signalers. Another possibility is that diffusion is again noise on the
realization of strategies: while behavior is optimized to be efficient
(be it due to learning, introspection or imitation), realization of
strategies is bound to be noisy due to confusability of similar
states.

We believe that all of the four mentioned interpretations are, on
first approximations, feasible conceptualizations of the \rdd, and
that it is a good thing to know of a working account of vague
signaling that sketches where fitness-based selection under
state-confusability will take is, abstracting away from the details of
actually playing the game, inheriting, imitating or otherwise
optimizing behavior in whatever particular way. It is a good thing to
know this on the macro-level, especially since there are also
micro-level accounts that nicely complement the picture. We turn to
one such next.

\subsection{Relation with previous accounts}
\label{sec:relat-with-prev}

@@
